---
title: "FeatureTable Summary Plots"
author: "Ryan M. Moore"
date: 2020-09-21
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FeatureTable Summary Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5.5
)
```

*Note: This vignette uses `s3` style functions.  `R6` methods are also avaiable.  See [this vignette](TODO) to see examples of how to use them.*

## Set up

First, import the libraries and data that we want to use.  We import `magrittr` so we can use the pipe, but you don't actually need it.

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(magrittr)
library(ggplot2)
library(featuretable)

data(lee)
```

## Basic summary plots

To start, let's plot the total count for each sample.  To do that, we use the `plot` function and pass in `fill = FALSE`.

```{r Plot with no fill}
plot(lee, fill = FALSE)
```


Next, let's see how to color the bars so that we can see how abundant individual ASVs are.  For that we just call the `plot` function with no additional arguments.

```{r Default plot}
plot(lee)
```

As you can see, a barplot is shown to display abundance info for ASVs in the samples.  By default, only the top 8 most abundant features are displayed.  The rest get grouped into the category `Other`.

Looks pretty good, but the default labels don't quite match up with our data.  Let's change them to fit.

```{r Plot with adjusted labels}
plot(lee, ylab = "Abundance", legend_title = "ASV")
```

Nice! There are a lot of ASVs, so this plot isn't the most interesting.  Let's collapse the features (a.k.a., the ASVs) by their taxanomic class first.  For that, we use the `collapse_features` function.

```{r Collapse by class}
lee %>% 
  collapse_features(Class) %>% 
  plot(ylab = "Abundance", legend_title = "Class")
```

Hmm, looks like the counts of Alpha and Gammaproteobacteria are pretty high, neat!

Of course, the raw counts don't give the best view of the data, so let's look at relative abundance instead.  Here's how we can do that.

```{r Adding relative abundance}
lee %>% 
  collapse_features(Class) %>% 
  map_samples(relative_abundance) %>% 
  plot(ylab = "Relative abundance", legend_title = "Class")
```

Whoa! Samples `BW1` and `R11BF` look quite different from the others.  Let's see what the data looks like if we group samples by `Char`.

Since there are only five levels in the `Char` variable, we can rotate the x axis labels so that they are parallel with the x axis.  To do that, use the `axis.text.x` parameter.  It works like the `axis.text.x` parameter in the `ggplot2::theme` functions: you pass it an `element_text` with the options you want.

And for fun, let's plot a few more of the features.  We do that by passing in `num_features = 12`.

```{r Show more features}
lee %>% 
  collapse_features(Class) %>% 
  collapse_samples(Char) %>% 
  map_samples(relative_abundance) %>% 
  plot(xlab = "Char", 
       ylab = "Relative abundance",
       legend_title = "Class",
       axis.text.x = element_text(angle = 0),
       num_features = 12)
```

Pretty nice looking plot there!

Just for fun, let's only keep features that are in the three classes at the bottom of the previous chart, `Alphaproteobacteria`, `Gammaproteobacteria`, and `Nitrosopumilales`, since they are highly abundant.  Then we can collapse those features by `Order`, to look at how the relative abundance of some of the most abundant Orders vary by `Char`.

```{r Keep by Class, then collapse by Order}
lee %>%
  keep_features(Class %in% c("Alphaproteobacteria", "Gammaproteobacteria", "Nitrosopumilales")) %>%
  collapse_features(Order) %>% 
  collapse_samples(Char) %>% 
  map_samples(relative_abundance) %>% 
  plot(xlab = "Char",
       ylab = "Relative abundance",
       legend_title = "Order",
       axis.text.x = element_text(angle = 0))
```

## Custom color palettes

### Picking a different color palette

The `plot` function in `FeatureTable` knows about five different color palettes:

- `kelly`
- `muted`
- `bright`
- `vibrant`
- `high contrast`

You can pick one of them using the `palette` parameter of the `plot` function.  Here is an example using the `muted` palette.

```{r Using muted palette}
# Save the data for use in subsequent plots.
lee_plot_data <- lee %>% 
  collapse_features(Class) %>% 
  collapse_samples(Char) %>% 
  map_samples(relative_abundance)

lee_plot_data %>% 
  collapse_features(Class) %>% 
  collapse_samples(Char) %>% 
  map_samples(relative_abundance) %>% 
  plot(xlab = "Char", 
       ylab = "Relative abundance",
       legend_title = "Class",
       axis.text.x = element_text(angle = 0),
       palette = "muted")
```

### Recycling colors

Sometimes, you want to dispaly data that has more features than colors in the included palette.  In these cases, the colors will be recycled.

```{r Recycling colors}
lee_plot_data %>% 
  plot(xlab = "Char", 
       ylab = "Relative abundance",
       legend_title = "Class",
       axis.text.x = element_text(angle = 0),
       num_features = 7,
       palette = "high contrast")
```

In this case, we asked to plot 7 features, but the `high contrast` palette only has 3 colors.  (The gray doesn't count as an official palette color.)  So you can see in the legend that the 3 colors of the palette get repeated to make up the difference.

Also worth noting is that the `Other` category will always be a gray color if you use one of the five built in palettes.

### Using your own palette

If you don't like any of the included palettes, then you can make your own with the `ggplot2` functions.  The `FeatureTable` plotting function returns a `ggplot` object, so you can add whatever you want to it.  To add a custom palette, you need to use one of the `scale_fill_*` functions, as the color is set using the `fill` parameter in the `aes` mapping function.  Something like this:

```{r Using discrete palette}
lee_plot_data %>% 
  plot(xlab = "Char", 
       ylab = "Relative abundance",
       axis.text.x = element_text(angle = 0)) +
  scale_fill_discrete(name = "Class")
```

Notice that I dropped the `legend_title` argument and passed `name = Class` to `scale_fill_discrete`.  This is because we're adding a new scale fill to the plot, so the old one gets overridden.

You may also notice that the `Other` category now is just the next color that `scale_fill_discrete` picks.  If you want to have the `Other` category be some gray color, the easiest thing would be to just use one of the built in palettes.  However, you could also make a custom scale like this:

```{r Using scale fill manual}
lee_plot_data %>% 
  plot(xlab = "Char", 
       ylab = "Relative abundance",
       legend_title = "Class",
       num_features = 3,
       axis.text.x = element_text(angle = 0)) +
  scale_fill_manual(name = "Class", values = c("#bbbbbb", "#B15928", "#FB9A99", "#1F78B4"))
```

Keep in mind that the colors you pass to the `values` argument of `scale_fill_manual` will go from top to bottom in the resulting legend.  So the first one should be the gray color you want for the `Other` category.  Finally, the `num_features` needs to be one less than the length of the color vector you pass to the `values` parameter, or it won't work properly.  

So you see, you *can* get the gray `Other` category with a custom color palette, but it is more work than using one of the default palettes.

